---
title: "Use Cases for Metabolomics Data Annotation Using MetaboAnnotation"
author: "Johannes Rainer, Andrea Vicini, Liesa Salzer, Jan Stanstrup, Josep M Badia,  \nSteffen Neumann, Michael Stravs, Vinicius Verri Hernandes, Laurent Gatto, Sebastian Gibb  \nand Michael Witting"
graphics: yes
output:
  rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Use Cases for Metabolomics Data Annotation Using MetaboAnnotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
  %\VignettePackage{MetaboAnnotationTutorials}
  %\VignetteDepends{xcms,Spectra,CompoundDb,pander,BiocStyle,MetaboAnnotation}
bibliography: references.bib
---

```{r biocstyle, echo = FALSE, message = FALSE }
library(BiocStyle)
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 7,
               dev = c("png"))
## BiocStyle::markdown()
```

This documents provides use cases describing the annotation of untarget
metabolomics and small compound datasets using functionality from the
`r Biocpkg("MetaboAnnotation")`, `r Biocpkg("MetaboCoreUtils")` and
`r Biocpkg("CompoundDb")` packages. All data used in this document is available
within the [github
repository](https://github.com/jorainer/annotation_ecosystem)
https://github.com/jorainer/annotation_ecosystem. Additional use cases and
examples can be found in the packages' vignettes and workshops/tutorials. See
section *Package repositories and additional tutorials* for links to the package
repositories and tutorials.

# Installation instructions

The R packages used in this document can be installed with the R code below:

```{r, eval = FALSE}
install.packages("BiocManager")
BiocManager::install(c("xcms", "pander"))
BiocManager::install(c("Spectra", "MsCoreUtils", "MetaboCoreUtils"))
BiocManager::install("RforMassSpectrometry/CompoundDb")
BiocManager::install("RforMassSpectrometry/MetaboAnnotation")
```

The source code for this document along with the test data can be downloaded
from the github repository https://github.com/jorainer/annotation_ecosystem with
the command (or alternatively downloading the zip archive directly from the
github page; note that [git-lfs](https://git-lfs.github.com/) needs to be
installed for the command below to work).

```
git clone https://github.com/jorainer/annotation_ecosystem
```

# $MS^1$ annotation

In this example we annotate chromatographic peaks identified in a test sample
provided within this repository first based only on their *m/z* values and then
using their *m/z* and retention times against previously determined values for a
set of pure standards. The test data represents LC-MS/MS measurements of a set
of metabolite standards. In brief, aliquots for pure standards of 15 metabolites
were solved in water and injected into a LC-MS system. Details are provided in
the Material and Methods section in the main text.

Below we load the data which is distributed within the
`MetaboAnnotationTutorials` package and perform the chromatographic peak
detection using the `r Biocpkg("xcms")` package.

```{r}
library(xcms)

fl <- system.file("mzML", "HighIS_Mix07_CE20_POS.mzML",
                  package = "MetaboAnnotationTutorials")

std_data <- readMSData(fl, mode = "onDisk")
```

We next perform the chromatographic peak detection using the *centWave*
algorithm followed by a *peak refinement* to reduce the number of peak detection
artifacts, such as split-peaks or overlapping peaks (see the *LC-MS/MS data
analysis with xcms* vignette from the `xcms` package for details on the
chromatographic peak detection settings).

```{r}
#' Peak detection
cwp <- CentWaveParam(snthresh = 10, prefilter = c(3, 4000), ppm = 40,
                     peakwidth = c(2, 8), integrate = 2)
std_data <- findChromPeaks(std_data, param = cwp)
#' Peak refinement
std_data <- refineChromPeaks(
    std_data, MergeNeighboringPeaksParam(expandRt = 3))
```

In total `r nrow(chromPeaks(std_data))` chromatographic peaks have been
identified. This is even more than would be expected since the sample represents
a mixture of 15 pure standards solved in water. Below we display the first 6 of
the detected chromatographic peaks.

```{r}
head(chromPeaks(std_data))
```

We next annotate these chromatographic peaks using only their *m/z* values. As a
reference we load a `CompDb` database containing all compounds from the Human
Metabolom Database [@wishartHMDBHumanMetabolome2018] version 5.0 which is also
provided within the `MetaboAnnotationTutorials` package. Such annotation
resources can be created with the `r Biocpkg("CompoundDb")` package (see the
[package documentation](https://RforMassSpectrometry.github.io/CompoundDb) for
more information).

```{r}
library(CompoundDb)

library(piggyback)

#' Download the SQLite database from github.
pb_download("CompDb.Hsapiens.HMDB.5.0.sqlite",
            repo = "jorainer/MetaboAnnotationTutorials",
            tag = "2021-11-02",
            dest = tempdir())

fl <- file.path(tempdir(), "CompDb.Hsapiens.HMDB.5.0.sqlite")

#' Load a CompDb database with compound annotation from HMDB
cdb <- CompDb(fl)
cdb
```

**Note**: while we are here annotating chromatographic peaks from an `xcms`
result object, it would also be possible to annotate LC-MS features using the
same approach by simply exchanging the `chromPeaks` call with
`featureDefinitions`. In addition, the annotation could also be performed on a
`SummarizedExperiment` representing the preprocessing results which can be
generated from an `xcms` result object using the `quantify` method.


## Annotation using *m/z* values

We next extract all peaks from the `xcms` result object and annotate them using
the `matchMz` function against all compounds from HMDB (the compound annotations
can be extracted from the `CompDb` database with the `compounds` function). The
reference database does not provide *m/z* values but only exact masses for each
compound. Thus, in order to annotate our chromatographic peaks' *m/z* values,
*m/z* values for expected ions of the compounds need to be calculated
first. With `Mass2MzParam` we can configure the parameters for the
annotation. Parameter `adducts` allows to specify the adducts (ions) we expect
to see mostly in the data set. For our example we assume to see mostly `[M+H]+`
and `[M+Na]+` positive ions. With parameters `tolerance` and `ppm` we can
finally define also the absolute and/or *m/z* relative acceptable difference,
respectively, for *m/z* values to be considered matching.

```{r}
library(MetaboAnnotation)

#' Get the chromatographic peaks and add a column with the peak_id
pks <- data.frame(peak_id = rownames(chromPeaks(std_data)),
                  chromPeaks(std_data))

#' Set parameters for the m/z-based annotation
param <- Mass2MzParam(adducts = c("[M+H]+", "[M+Na]+"),
                      tolerance = 0, ppm = 10)

#' Perform the matching
pks_match <- matchMz(
    pks,
    compounds(cdb, c("compound_id", "exactmass", "formula", "name")),
    param = param)

pks_match
```

The result object handles the potential many-to-many matching between
chromatographic peaks (*query*) and reference (*target*) and contains all
information from both the query and the target object along with the *score* for
the match (in this case the difference between the *m/z* values). Below we list
all available information in the result object.

```{r}
colnames(pks_match)
```

The n:m mapping is represented in the `matches` data frame. It maps each
annotated query element to all matching target elements. Below we list the first
6 rows from that table. Column `"score"` provides the difference between the
query and target *m/z* values (in Da).

```{r}
head(matches(pks_match))
```

For users it's however more convenient to extract annotation information with
the `matchedData` function, that will return a data frame with all
annotations. If query elements are annotated to more than one target element
they will be listed in as many rows as there are matches. Not matching query
elements are represented with a single row in the result table, containing `NA`
for all annotations. Below we extract all annotations, restricting to some of
the annotation columns.

```{r}
matchedData(pks_match, c("peak_id", "target_name",
                         "target_formula", "adduct"))
```

As we can see, the number of rows of this data frame is much larger than the
number of chromatographic peaks. Most peaks are in fact annotated to several
possible compounds based on their *m/z*. The average number of compounds to
which a chromatographic peak in this data is annotated is calculated below.

```{r}
matches(pks_match)[, 1] |>
table() |>
mean()
```

Annotations in the result object are organized by query, annotations for
individual query elements can thus be accessed very easily. Below we list all
annotations for the first chromatographic peak.

```{r}
matchedData(pks_match[1], c("peak_id", "target_name",
                            "target_formula", "adduct"))
```

All compounds have thus the exact same chemical formula and hence, using only
the *m/z* value, we can not discriminate between them or determine which would
be the correct annotation. Considering also the retention times would allow a
better annotation and increase its confidence.


## Annotation using reference *m/z* and retention times

For LC-MS experiments compounds are first chromatographically separated based on
some property of the compound (other than their mass). Thus, also the retention
times contain information that can improve the annotation. In this section we
use information from a previous experiment in which pure standards were measured
with the same LC-MS setup used to generate the present test data and in which
the retention time as well as the ion (adduct) and its *m/z* value were
determined. These reference values were thus empirically determined on the same
instruments with the same setup and should thus help to get higher confidence
annotations.

Below we load the reference retention time and *m/z* data for some of the
measured standards.

```{r}
fl <- system.file("txt", "std_ions.txt",
                  package = "MetaboAnnotationTutorials")
std_ions <- read.table(fl, sep = "\t", header = TRUE)
head(std_ions)
```

We can now annotate the chromatographic peaks using this reference data. We
configure the matching using the `MzRtParam`, which allows to annotate query
elements based on *m/z* and retention times. With `ppm` we define again the
*m/z*-relative acceptable difference of the *m/z* values and with `toleranceRt`
the absolute acceptable difference in measured retention times. The parameters
`mzColname` and `rtColname` allow us to define the names of the columns in the
query and target objects containing the *m/z* values and retention time values
for the matching. Settings for `ppm` and `toleranceRt` should be adapted to the
experimental setup, retention time difference could e.g. be larger for
HILIC-based chromatographic separation. Below we use a difference of 7 seconds
(since retention times in `std_ions` and in the `xcms` result are reported in
seconds).

```{r}
param <- MzRtParam(ppm = 10, toleranceRt = 7)
pks_match <- matchMz(pks, std_ions, param = param,
                     mzColname = c("mz", "ion_mz"),
                     rtColname = c("rt", "ion_rt"))
pks_match
```

In addition to supplying the reference (target) data as a `data.frame` we could
also directly use an `IonDb`. This type of annotation object extends the above
used `CompDb` database providing in addition to general compound annotations
also information on measured ions including their retention times and *m/z*
values. It is thus designed to contain reference annotation databases for
specific LC-MS setups and MS instrumentation used in a lab. Below we create such
an `IonDb` database (which will be saved in the present example to a temporary
file) from the HMDB `CompDb` database and add our ion annotations to it.

```{r}
#' Create an IonDb filling it with data from cdb.
idb <- IonDb(tempfile(), cdb)

#' Insert measured m/z and retention times for ions
idb <- insertIon(idb, std_ions)
idb
```

The advantage of such an annotation database over individual (text) file-based
annotation approaches is obvious: along with the annotation data also metadata
can be saved and the database can be versioned, which is crucial for
reproducible workflows. Also, the database's SQLite file is portable and can be
distributed and used on different computers. In addition, it would also be
possible to use a MySQL-based database system which would enable centralized
annotation resources in a laboratory. More information on the creation and usage
of annotation resources with the `CompoundDb` package is provided in its package
vignettes which are also available
[here](https://RforMassSpectrometry.github.io/CompoundDb/).

We can now annotate our peaks against that database using the same call as
above, but providing the `idb` as `target` parameter.

```{r}
pks_match <- matchMz(
    pks, ions(idb, c("compound_id", "ion_adduct", "ion_mz", "ion_rt", "name")),
    param = param, mzColname = c("mz", "ion_mz"),
    rtColname = c("rt", "ion_rt"))
pks_match
```

In total `r length(whichQuery(pks_match))` chromatographic peaks were annotated
to an ion of one of the standards. We can identify the chromatographic peaks
with available annotation using the `whichQuery` function, that returns their
index in the query. Below we thus also subset the matched results to only those
chromatographics peaks with an annotation.

```{r}
whichQuery(pks_match)

pks_match <- pks_match[whichQuery(pks_match)]
```

We next build a result table for the successfully annotated chromatographic
peaks.

```{r, results = "asis"}
library(pander)
pandoc.table(as.data.frame(
    matchedData(pks_match, c("peak_id", "target_name", "score", "score_rt"))),
    style = "rmarkdown", split.tables = Inf)
```

Columns `"score"` and `"score_rt"` contain the difference between the query and
the annotated target's *m/z* (in Da) and retention time (in seconds),
respectively.


## Annotation using reference *m/z* and retention indices

Retention time is a valuable orthogonal information for metabolite
identification, but chromatographic conditions are far from being
standardized. Even when using nominally the same chromatographic system, meaning
column and eluents, differences in retention times arise due to dead volumes,
gradient delay volumes, etc. In GC-MS retention time indexing has been used to
normalize for drifts in retention times by converting them into retention
indices. A similar approach has been recently described for LC-MS
[@stoffelNAlkylpyridiniumSulfonatesRetention2021] using N-Alkyl-pyridinium
sulfonates as retention indexing substances. Their use allows to convert
retention times into retention indices and to compare retention information
across different LC-MS systems.

We first load data from an untargeted metabolomic experiments. This data has
been completely processed. Each feature is identified by a unique name.

```{r}
library(MetaboCoreUtils)

fl <- system.file("txt", "rti_ms_pos_features.txt",
                  package = "MetaboAnnotationTutorials")
ms1_features <- read.table(fl, header = TRUE, sep = "\t")
head(ms1_features)
```

Likewise we load the information on the N-Alkyl-pyridinium sulfonates. The
retention index is defined as the number of carbons in the alkyl chain times
100.

```{r}
fl <- system.file("txt", "rti_ms_pos_naps.txt",
                  package = "MetaboAnnotationTutorials")
ms1_naps <- read.table(fl, header = TRUE, sep = "\t")
head(ms1_naps)
```

The relationship between the retention time and retention index can be plotted.

```{r}
plot(ms1_naps)
```

We can use the N-Alkly-pyridinium sulfonates to convert the retention times of
the measured features using the `indexRtime` function. By default linear
interpolation is used for conversion, but other (custom) functions can be
defined if required.

```{r}
ms1_features$rindex <- indexRtime(ms1_features$rtime, ms1_naps)
```

This data can then be annotated using reference databases. Two different
databases are used. One was obtained on a UHPLC column and one on a HPLC column
of the same column chemistry. The untargeted data was obtained from a HPLC
column, but using a different LC hardware, resulting in different absolute
retention times.

```{r}
fl <- system.file("txt", "rti_db_hplc.txt",
                  package = "MetaboAnnotationTutorials")
db_hplc <- read.table(fl, header = TRUE, sep = "\t",
                      quote = "", comment.char = "")

fl <- system.file("txt", "rti_db_uhplc.txt",
                  package = "MetaboAnnotationTutorials")
db_uhplc <- read.table(fl, header = TRUE, sep = "\t",
                      quote = "", comment.char = "")
```

Next we can define the `Mass2MzRtParam` similar to above. We will use an
absolute error of 10 retention index units and use `[M+H]+ `and `[M+Na]+`
adducts.

```{r}
match_param <- Mass2MzRtParam(adducts = c("[M+H]+", "[M+Na]+"),
                              tolerance = 0.005,
                              toleranceRt = 10)
```

Matching can now be performed. Important is that the column with the retention
index needs to be defined using `rtColname`. Matching is performed against both
databases and results are then filtered to only contain features with at least
one match.

```{r}
matches_hplc <- matchMz(ms1_features,
                        db_hplc,
                        match_param,
                        massColname = "exact.mass",
                        mzColname = "mz",
                        rtColname = "rindex")

matches_hplc <- matches_hplc[whichQuery(matches_hplc)]

matches_uhplc <- matchMz(ms1_features,
                         db_uhplc,
                         match_param,
                         massColname = "exact.mass",
                         mzColname = "mz",
                         rtColname = "rindex")

matches_uhplc <- matches_uhplc[whichQuery(matches_uhplc)]

```

We next build a result table for the successfully annotated chromatographic
peaks. First for the results matching with the HPLC database.

```{r, results = "asis"}
library(pander)
pandoc.table(head(as.data.frame(
  matchedData(matches_hplc, c("Name", "mz", "rindex", "target_name",
                              "adduct", "score", "score_rt")))),
  style = "rmarkdown", split.tables = Inf)
```

Then for matching with the UPLC database.

```{r, results = "asis"}
library(pander)
pandoc.table(head(as.data.frame(
  matchedData(matches_uhplc, c("Name", "mz", "rindex", "target_name",
                               "adduct", "score", "score_rt")))),
  style = "rmarkdown", split.tables = Inf)
```


# $MS^2$ annotation

In addition to $MS^1$ annotation described in the previous section, it is also
possible to annotate features (or chromatographic peaks) by comparing
experimentally measured $MS^2$ (MS/MS) spectra against reference spectra. Such
annotation can be performed using the `matchSpectra` from the `MetaboAnnotation`
package that takes `Spectra` objects (with $MS^2$ spectra) as query and target
parameters. Such `Spectra` objects can be either imported from a variety of
input formats, including MGF, MSP, mzML, mzXML or netCDF files, extracted from
`xcms` result objects or even retrieved from databases such as *MassBank* (see
also the tutorials from
[SpectraTutorials](https://jorainer.github.io/SpectraTutorials/) for more
examples and use cases).

To illustrate this, we will extract in this section $MS^2$ spectra for the
annotated chromatographic peaks from the previous section and compare them
against reference $MS^2$ spectra from HMDB. Annotation reliability can be
increased if, in addition to *m/z* and retention time, also the $MS^2$ spectra
for a chromatographic peak or a feature matches the reference spectrum from a
pure standard.

We first extract all $MS^2$ spectra for the annotated chromatographic peaks
using the `chromPeakSpectra` function from the `xcms` result object. This will
return all $MS^2$ spectra with the precursor *m/z* and retention time within the
*m/z* ranges and retention time ranges of the detected chromatographic
peaks. For an `xcms` result object containing also correspondence results
(i.e. grouped chromatographic peaks across samples, also referred to as
*features*) the `featureSpectra` function could be used instead.

```{r}
std_spectra <- chromPeakSpectra(std_data, return.type = "Spectra",
                                peaks = pks_match$peak_id)
std_spectra
```

For some of the chromatographic peaks we have more than one $MS^2$ spectrum, but
in total we have $MS^2$ spectra only for `r length(unique(std_spectra$peak_id))`
chromatographic peaks. Below we count the number of spectra per peak.

```{r}
table(std_spectra$peak_id)
```

Before matching the spectra we are *cleaning* our experimental spectra removing
peaks with an intensity lower than 5% of a spectra's highest peak intensity and
removing spectra with less than 3 peaks.

```{r}
#' Define a function to remove low intensity peaks
low_int <- function(x, ...) {
    x > max(x, na.rm = TRUE) * 0.05
}
#' Remove peaks with an intensity below 5% of BPI
std_spectra <- filterIntensity(std_spectra, intensity = low_int)

#' Remove peaks with less than 3 peaks
std_spectra <- std_spectra[lengths(std_spectra) > 2]
std_spectra
```

In addition, we scale the peak intensities within each spectrum to values
between 0 and 100. This is mostly for visualization reason, since most spectra
similarity scoring algorithms are independent of absolute peak intensities.

```{r}
#' Define a function to *scale* the intensities
scale_int <- function(x, ...) {
    maxint <- max(x[, "intensity"], na.rm = TRUE)
    x[, "intensity"] <- 100 * x[, "intensity"] / maxint
    x
}
#' *Apply* the function to the data
std_spectra <- addProcessing(std_spectra, scale_int)
```

Below we match now these *cleaned* spectra against the reference spectra from
HMDB. We configure the matching using the `CompareSpectraParam` allowing *m/z*
differences of 50ppm between peaks from a query and a target spectrum. Since
HMDB does not provide precursor *m/z* we in addition have to disable the
additional matching filter (i.e. set `requirePrecursor = FALSE`), which has
however a negative impact on the performance of the comparisons. By default,
spectra similarities (calculated with the normalized dot product) higher than
0.7 will be considered matching.

```{r}
spectra_match <- matchSpectra(
    std_spectra, Spectra(cdb),
    param = CompareSpectraParam(ppm = 50, requirePrecursor = FALSE))
spectra_match
```

Thus, in total `r length(whichQuery(spectra_match))` of the query spectra were
matched with `r length(whichTarget(spectra_match))`. Below we create a result
table for this matching.

```{r, results = "asis"}
res <- spectraData(spectra_match,
                   c("peak_id", "target_compound_id", "score"))
cmp_id <- unique(res$target_compound_id)
cmp_id <- cmp_id[!is.na(cmp_id)]
cmps <- compounds(cdb, filter = ~ compound_id == cmp_id, "name")
rownames(cmps) <- cmps$compound_id

res$name <- cmps[res$target_compound_id, "name"]
rownames(res) <- NULL
pandoc.table(as.data.frame(res),
    style = "rmarkdown", split.tables = Inf)

```

Matching results can also be very easily visualized with the `plotSpectraMirror`
function. Below we show the results for the first peak (Caffeine). The upper
spectrum represents the query, the lower the target (reference) spectrum.

```{r}
plotSpectraMirror(spectra_match[1])
```

The results seem to be reasonable. For the second peak (third spectrum) the
match was based however on a single peak.

```{r}
plotSpectraMirror(spectra_match[3])
```

Note that we could now also add some of the (cleaned and annotated) $MS^2$
spectra from the present test set, to the `IonDb` using the `insertSpectra`
function. This enables generation of lab-internal reference databases including
also MS/MS spectra.

# Package repositories and additional tutorials

- Tutorial with additional examples and explanations for $MS^2$-based
  annotations: https://jorainer.github.io/SpectraTutorials/
- Repository of the `MsCoreUtils` package:
  https://rformassspectrometry.github.io/MsCoreUtils/
- Repository of the `MetaboCoreUtils` package:
  https://rformassspectrometry.github.io/MetaboCoreUtils/
- Repository of the `Spectra` package:
  https://rformassspectrometry.github.io/Spectra/
- Repository of the `MetaboAnnotation` package:
  https://rformassspectrometry.github.io/MetaboAnnotation/
- Repository of the `CompoundDb` package:
  https://rformassspectrometry.github.io/CompoundDb/


# Session information {-}

The R version and versions of used packages are listed below.

```{r}
sessionInfo()
```

# References
